---
title: "RBIF111 Homework 1"
author: "Rita Pecuch"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Objective: Create a function that calculates the mean of a numeric vector starting from the first element to the last. Then compare this to the results of the mean() R function.

Function definition:

```{r mean function}
calculate_mean <- function(vector, na.rm=F){
  # Remove NA values if needed
  if (na.rm){
    vector <- vector[!is.na(vector)]
  }
  # Calculate sum of all observations
  total <- 0
  for(x in vector){
      total <- total + x
  }
  
  # Divide total by number of observations
  mean <- total / length(vector)
  return(mean)
}
```

Comparing function results with mean() R function:

```{r comparing mean functions}
test_vector <- c(1,2,3,4,5,6,7,11,100)
# R function
mean(test_vector)
# Custom function
calculate_mean(test_vector)

# Add NA value
test_vector_na <- c(test_vector, NA)
mean(test_vector_na)
calculate_mean(test_vector_na)

# Add NA value and use na.rm=T
mean(test_vector_na, na.rm=T)
calculate_mean(test_vector_na, na.rm=T)
```

## Question 2

Objective A: Execute and explain the results of the following piece of R code:


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
