---
title: "RBIF111 Homework 1"
author: "Rita Pecuch"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Objective: Create a function that calculates the mean of a numeric vector starting from the first element to the last. Then compare this to the results of the mean() R function.

Function definition:

```{r mean function}
calculate_mean <- function(vector, na.rm=F){
  # Remove NA values if needed
  if (na.rm){
    vector <- vector[!is.na(vector)]
  }
  # Calculate sum of all observations
  total <- 0
  for(x in vector){
      total <- total + x
  }
  
  # Divide total by number of observations
  mean <- total / length(vector)
  return(mean)
}
```

Comparing function results with mean() R function:

```{r comparing mean functions}
test_vector <- c(1,2,3,4,5,6,7,11,100)
# R function
mean(test_vector)
# Custom function
calculate_mean(test_vector)

# Add NA value
test_vector_na <- c(test_vector, NA)
mean(test_vector_na)
calculate_mean(test_vector_na)

# Add NA value and use na.rm=T
mean(test_vector_na, na.rm=T)
calculate_mean(test_vector_na, na.rm=T)
```

## Question 2

Objective A: Execute and explain the results of the following block of R code:

```{r}
mat <- matrix(1:9, nrow = 3, ncol = 3)
```
This line of code uses the matrix() function to generate a matrix with numerical data values 1-9. The matrix has 3 rows and 3 columns, and the data values fill in the first column, then the second, then the third. There is no need for recycling because the number of data values equals the dimensions of the matrix. Because the result of the matrix() function is being stored in a variable, no nothing is printed.

```{r}
which.max(mat)
```
This line of code uses the which.max() function to determine the location of the maximum of the matrix, which is position 9. The result is not being stored in a variable and thus is printed.

```
which.max(as.data.frame(mat))
```
TODO: come back to this
This line of code attempts to use the which.max() function on the matrix that has been coerced to a dataframe. An error message is printed because

```
mat[1:9,]
```
This line of code attempts to subset a matrix, but an error message is printed because the first argument is specifying rows 1-9. This matrix only has rows 1-3, as it is a 3 x 3 dimension matrix.

```
mat[,1:9]
```
This line of code attempts to subset a matrix, but an error message is printed because the second argument is specifying columns 1-9. This matrix only has columns 1-3, as it is a 3 x 3 dimension matrix.


Objective B: Execute and explain the results of the following block of R code:

```{r}
m <- matrix(c(1:5,NA,7,7,NA), nrow=3,ncol=3,byrow=T); print(m)
```
This line of code uses the matrix() function to generate a matrix with the following values: 1, 2, 3, 4, 5, NA, 7, 7, and NA. The matrix has 3 rows and 3 columns, and because byrow=T is specified the data values fill in the first row, then the second, then the third. There is no need for recycling because the number of data values equals the dimensions of the matrix. Nothing is printed when the result of a function is being stored in a variable, but the print() function is then called which prints the matrix.

```{r}
numbers <- NULL
```
This line of code assigns the value NULL to the variable called numbers. Nothing is printed because the value is being stored in a variable. When used with the below chunk of code, this line of code serves the purpose of initializing the numbers variable with an empty value.

```
for(i in 1:4){
  temp <- m[i,]
  numbers <- c(numbers, temp)
}
```
This block of code attempts to loop through the numbers in the range 1-4 and apply the following logic at each iteration:
- Subset a single row from matrix m with the row number matching the current iteration of the loop
- Add all of the values in the subsetted matrix (a vector) to the current values of the numbers variable.
However, an error message is printed on the final iteration of the loop because matrix m is a 3 x 3 matrix and does not have a row 4. Because the loop was able to successfully complete iterations 1-3, the current value of the numbers variable is c(1, 2, 3, 4, 5, NA, 7, 7, NA).

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
